
**the cholera model run with different parameters**

```{r}
#reading in 
library(readr)
library(dplyr)
library(readxl)
library(ggplot2)
cholinc <- read_excel("choleracases.xlsx")

cholinc <- cholinc %>% mutate ("time" = seq(1:82)) #adding time column 

```

```{r}
#vectorizing 
time <- cholinc$time
IncData <- cholinc$Cases
numPoints <- length(time)
```

```{r}
## Visualize the data:
library(ggplot2)
ggplot(cholinc, aes(x = time, y = IncData)) + geom_point() +
  geom_line(data = cholinc, aes(x = time, y = IncData))

```

```{r}
#running the model
SIRD <- function(time, state, theta) {
  ## Parameters:
  delta <- case_when(
    time >= 0 & time <= 9 ~ theta["decaylow"],
    time >= 20 & time <= 31 ~ theta["decaylow"],
    time >= 63 & time <= 69 ~ theta["decaylow"], #decay rates from Codeco paper
    TRUE ~ theta["decayhigh"]
  )
  gamma <- 1 / theta["Duration"]
  psi <- theta["psi"]
  kappa <-  theta["kappa"]
  mu <- theta["mu"]
  epsilon <- theta["epsilon"]
  
  ## States:
  S <- state["S"]
  I <- state["I"]
  R <- state["R"]
  D <- state["D"]
  W <- state["W"]
  
  
  ## ODEs:
  dS <- -(W/(kappa+W)) * psi * S 
  dI <- (W/(kappa+W)) * psi * S - (gamma + mu) * I
  dR <- gamma * I
  dD <- mu*I
  dW <- epsilon*I - delta*W
 
  
  return(list(c(dS, dI, dR, dD, dW)))
}
```

#derived incidence
  incidence <- ((W/(kappa+W)) * psi * S)
  
#derived CFR
 CFR <- mu*I / (mu *I + gamma * I)


```{r}
library(deSolve)

#theta from MCMC estimation 1

thetamcmc1 <- c(decayhigh = 1/3, #from cholera dynamics for public health practitioners 
                decaylow = 1/41, 
                Duration = .29, #2/7th of a week, rate of leaving is inverted 
                psi = .00624, #MCMC 
                mu = .012, #death rate (CFR)
                epsilon = 117.64, #mcmc vibrio/ml*person*week
                kappa = 1033003) #MCMC, vibrio/ml at which 50% infection is achieved (lower than literature)


#initial state (ouest population)
initState <- c(S = (4*10^6)-1, I = 1, R=0, D = 0, W = 4000)

#trajectory of model 1 using MCMC 1 values and literature values
trajmcmc1 <- data.frame(ode(y=initState, times=seq(from = 0 , to = 52*4, by = .1), func=SIRD,parms=thetamcmc1, method = "ode45"))  %>% 
  mutate ('R0' = thetamcmc1[['Duration']] * thetamcmc1[['psi']] * (W/(thetamcmc1[['kappa']] + W))) #coding for variable R0

plotR0 <- trajmcmc1 %>% 
  ggplot(aes(x = time, y = data, color = State)) +
  geom_line(aes(y = R0, col = "R0"), linewidth = 1.2)
plotR0

plotmcmc1 <- trajmcmc1 %>% 
  ggplot(aes(x = time, y = data, color = State)) +
  geom_line(aes(y = D, col = "D"), linewidth = 1.2) +
  geom_line(aes(y = S, col = "S"), linewidth = 1.2) +
  geom_line(aes(y = I, col = "I"), linewidth = 1.2) +
  geom_line(aes(y = R, col = "R"), linewidth = 1.2)
plotmcmc1


plotmcmc1.I <- trajmcmc1 %>% 
  ggplot(aes(x = time, y = data, color = State)) +
  geom_line(aes(y = I, col = "I"), linewidth = 1.2)
plotmcmc1.I
```

```{r}
#solving for easy R0

thetamcmc1 <- c(decayhigh = 1/3, #from cholera dynamics for public health practitioners 
                decaylow = 1/41, 
                Duration = .29, #2/7th of a week, rate of leaving is inverted 
                psi = .00624, #MCMC 
                mu = .012, #death rate (CFR)
                epsilon = 117.64, #mcmc vibrio/ml*person*week
                kappa = 1033003) #MCMC, vibrio/ml at which 50% infection is achieved (lower than literature)



R0e <- .00624 * 4.28 * (1033003/(1033003+1033003))
R0e
```

```{r}
#parameters from redone MCMC (beimnets code)

thetamcmc2 <- c(decayhigh = 1/3, decaylow = 1/41, Duration = 0.29605, 
           psi = 0.02634, kappa = 880109.75, mu = .012, epsilon = 35)


#initial state (ouest population)
initState <- c(S = (4*10^6)-1, I = 1, R=0, D = 0, W = 4000)


trajmcmc2 <- data.frame(ode(y=initState, times=seq(from = 0 , to = 52*4, by = 0.1), func=SIRD, parms=thetamcmc2, method = "ode45"))  



plot1 <- traj1 %>% 
  ggplot(aes(x = time, y = data, color = State)) +
  geom_line(aes(y = S, col = "S"), linewidth = 1.2) +
  geom_line(aes(y = D, col = "D"), linewidth = 1.2) +
  geom_line(aes(y = I, col = "I"), linewidth = 1.2) +
  geom_line(aes(y = R, col = "R"), linewidth = 1.2)
plot1

plot2 <- traj1 %>% 
  ggplot(aes(x = time, y = data, color = State)) +
  geom_line(aes(y = R, col = "R"), linewidth = 1.2)
plot2


```



```{r}

theta2 <- c(decayhigh = 1/3, decaylow = 1/41, Duration = 2/7, 
            psi = 10^-1.5, kappa = 10^5, mu = .012, epsilon = 35)

initState <- c(S = 10^6, I = 1, R = 0, D = 0, W = 0, incidence = 0, CFR = 0)

traj <- data.frame(ode(y=initState, times=seq(from = 0 , to = 82, by = 1), func=SIRD,
                       parms=theta2, method = "ode45")) 



```







initState <- c(S=2e6-1, I = 1, R=0, D =0, W = 0, incidence = 0)
traj <- data.frame(ode(y=initState, times=seq(from = 0 , to = 82, by = 1), func=SIRD,
                       parms=theta1, method = "ode45"))

plot <- traj %>% 
  ggplot(aes(x = time, y = data, color = State)) +
  geom_line(aes(y = S, col = "S"), linewidth = 1.2) +
  geom_line(aes(y = D, col = "D"), linewidth = 1.2) +
  geom_line(aes(y = I, col = "I"), linewidth = 1.2) +
  geom_line(aes(y = R, col = "R"), linewidth = 1.2)
plot

Infectedplot <- traj %>% 
  ggplot(aes(x = time, y = data, color = State)) +
geom_line(aes(y = I, col = "I"), linewidth = 1.2) 

Infectedplot

Incidenceplot <- traj %>% 
  ggplot(aes(x = time, y = data, color = State)) +
geom_line(aes(y = incidence, col = "incidence"), linewidth = 1.2) 

Incidenceplot


## Prior function:
logPrior <- function(theta) {
  # Prior on R0:
  logPriorkappa <- dunif(theta[["kappa"]], min = 10^5, max = 10^6, log = TRUE)
  
  return(logPriorkappa)
}

## Likelihood function for a single data point:
pointLogLike <- function(i, IncData, IncModel) {
  # Incidence is observed through a Poisson process.
  poissonLike <- dpois(x=IncData[i], lambda=IncModel[i], log=TRUE)
  if (is.na(poissonLike)) {
    return( -Inf )
  } else {
    return( poissonLike )
  }
}

## Likelihood function for all data points:
trajLogLike <- function(time, IncData, theta, initState) {
  trajModel <- data.frame(ode(y=initState, times=time, func=SIRD,
                              parms=theta, method = "ode45"))

  IncModel <-   trajModel$I
  
  logLike <- 0
  for (i in 1:length(time)) {
    logLike <- logLike + pointLogLike(i, IncData, IncModel)
  }
  return(logLike)
}

## Posterior function:
logPosterior <- function(time, IncData, theta, initState) {
  ## Calculate the log prior (logPrior) for the vector of model
  ## parameters (theta).
  logPrior <- logPrior(theta)
  
  ## Calculate the log likelihood (logLike) of the data given theta, the
  ## incidence data (IncData), and the initial values of the state
  ## variables (initState).
  logLike <- trajLogLike(time, IncData, theta, initState)
  
  logPosterior <- logPrior + logLike
  return(logPosterior)
}

logPosteriorMH <- function(MHparams) {
  return(logPosterior(time, IncData,
                      theta = c(MHparams, decayhigh = 1/3, decaylow = 1/41, Duration = 2/7, 
                                psi = 10^-1.9, mu = .012, epsilon = 35),
                      initState = c(S=2500000 - 1, I=1, R=0, D = 0, W = 0)))
}

## test

MHparams <- c(kappa = 10^6)

logPosteriorMH(MHparams)



#############HASTINGS ALGO####################################

library(truncnorm)

mcmcMH <- function(posterior, initTheta, proposalSD, numIterations) {
  
  # Evaluate the function "posterior" at "initTheta", and assign to a
  # variable called posteriorThetaCurrent.
  posteriorThetaCurrent <- posterior(initTheta)
  
  # Initialise variables to store the current value of theta, the
  # vector of sample values, and the number of accepted proposals.
  thetaCurrent <- initTheta
  samples <- initTheta
  accepted <- 0
  
  # Run the MCMC algorithm for numIterations interations.
  for (i in 1:numIterations) {
    
    # Draw a new theta from a Gaussian proposal distribution and
    # assign this to a variable called thetaProposed.
    thetaProposed <- rnorm(n = length(thetaCurrent),
                           mean = thetaCurrent, sd = proposalSD)
    
    # Assign names to the thetaProposed vector.
    names(thetaProposed) <- names(thetaCurrent)
    
    # Evaluate the log) posterior function at the proposed theta
    # value and assign to a variable called
    # posteriorThetaProposed.
    posteriorThetaProposed <- posterior(thetaProposed)
    
    # Compute the Metropolis-Hastings (log) acceptance
    # probability and assign to a variable called
    # logAcceptance.
    logAcceptance <- posteriorThetaProposed -posteriorThetaCurrent
    
    # Draw a random number uniformly-distributed between 0 and 1
    # using "runif" and assign to a variable called randNum.
    randNum <- runif(n = 1, min = 0, max = 1)
    
    # Use the random number and the acceptance probability to
    # determine if thetaProposed will be accepted.
    if (randNum < exp(logAcceptance)) {
      
      # If accepted, change the current value of theta to the
      # proposed value of theta.
      thetaCurrent <- thetaProposed
      
      # And update the current value of the posterior
      # function.
      posteriorThetaCurrent <- posteriorThetaProposed
      
      # And update number of accepted proposals.
      accepted <- accepted + 1
    }
    
    # Add the current theta to the vector of samples.
    samples <- c(samples, thetaCurrent)
    
    # Print the current state of chain and acceptance rate.
    cat("iteration:", i, "chain:", thetaCurrent,
        "acceptance rate:", accepted / i, "\n")
  }
  return(samples)
}


mcmcTrace <- mcmcMH(posterior = logPosteriorMH, # posterior distribution
                    initTheta = c(kappa = 500000, ), # intial parameter guess
                    proposalSD = c(500000*.01), # standard deviations of #0.075
                    # parameters for Gaussian proposal distribution
                    numIterations = 2000)

trace <- matrix(mcmcTrace, ncol = 1, byrow = T)



# Use the package "coda" to convert the trace into this format:
library(coda)
trace.1st <- mcmc(trace)
plot(trace.1st)
summary(trace.1st)











